Leadership across cultures and contexts: an exploratory analysis

### 2020-04-29 leadership_across_cultures_contexts.Rmd
Working on MS. Reorganized the main intro section a bit. Added to Study Aims.

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, competing leaders, pressures from rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural evolutionary processes [@henrich_big_2015].

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have been relatively under-investigated. Current outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation at different levels of social organizations and the costs and benefits leaders and followers incur across distinct leadership systems.

Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with specific rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

@@

Leaders emerge in all human societies. That is, within every population there are particular contexts in which one individual maintains a disproportionate level of influence in the group, functions to facilitate group goals, and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

@@

We then used several methods to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing 95% CIs as above.

To supplement results from the aforementioned methods, we also performed textual analyses...OTHER METHODS. NEED TO DO.

>>>>

Across human societies, the influence and agendas of leaders stands to substantially shape group structure and group outcomes. The nature of group structure, group goals, and the demands of leadership can favor particular qualities of leaders as well as the costs and benefits to individulas. Cultural norms, capacities of competing leaders, inter-group dynamics, and followers preferences can also impact leader agendas. The interactions between leader strategies, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural and biological evolutionary processes [@henrich_big_2015].

Leadership has been the focus of much scientific work across psychology, sociology, political science, and managerial studies. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020].

There is now well established consensus within behavioral and evolutionary human sciences on the importance of data from the full range of human cultural diversity and systemic biases in psychological sciences are well known [i.e., @henrich_weirdest_2010]. However, a categorization of populations as WEIRD or non-WEIRD and attempting to conceptualize "non-WEIRD" social dynamics to learn about human psychology is a reductionist correction and arguably only a marginal scientific improvement from exclusively studying WEIRD populations. Extending the study of leadership to include evolutionary theory and data from diverse, small-scale and traditional societies (i.e., non-WEIRD) can facilitate more generalizable theories and offer applications to shape organizational structure across populations (i.e., WEIRD settings) in ways which are more consistent and less mismatched with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019; @van_vugt_selected_2010].

Much of leadership studies, including evolutionary approaches, have focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and levels of social organization, such as among kin groups, economic groups, and sub-groupings of individuals within residential communities. Lastly, while the benefits leaders accrue for their services have received substantial attention -- fitness benefits in particular -- the costs of leadership for both leaders and followers have been relatively under-investigated. Outstanding questions for evolutionary leadership scholars include the degree to which the nature of leadership is associated with population-level variation and in different social contexts and identifying the associated benefits *and* costs both leaders and followers incur.

<!-- THIS PARAGRAPH CAN BE INCORPORATED ABOVE -->
Evolutionary scholars typically define leaders as individuals who maintain disproportionate influence over decision-making within groups [@von_rueden_leadership_2015], whatever the group may be. Leadership roles are also often associated with social sanctioned rights and responsibilities. Not surprisingly, the functions leaders serve then are often closely linked to group contexts and the salient challenges groups face. Certain types of groups, such as kin groups, economic groups, or political groups, may tend to favor leaders with specific qualities, however there can be substantial variation in the traits of leaders within these groups as well as between them.

@@

Within every population there are particular contexts in which one or a few individuals maintain a disproportionate level of influence in the group, function to facilitate group goals, and provides services to the group. Adopting the broad definition of leadership given above, parents can also be defined as leaders of nuclear families. Strong community leadership, however, is extremely common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @von_rueden_leadership_2015].

@@

Our second goal was to extract higher-order dimensions of leader qualities and functions from our researcher coded variables. We used several methods to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, non-negative matrix factorization (NMF), and hierarchical cluster analysis. Logistic PCA generalizes Pearson's original formulation of PCA to find a projection of the natural parameters of the saturated model that minimizes the Bernoulli deviance [@logisticPCA].

@@

With an interpretable method of dimensionality reduciton, we aimed to predict variation in our "feature" dimensions from populaiton-level variation, such as subsistence type, region, and group structure.

@@

Our third goal was to determine which leader and follower benefits and costs were most commonly mentioned in the ethnographic record, computing 95% CIs as above and to estimate which group structure types were associated with the most overall benefits and costs for leaders.

To supplement results from the aforementioned methods, we also performed textual analyses and explored varition measures of biases stemming from the ethnographic record that may have impacted results.
